---
title: "Code Review: h2 and rg estimation using Haseman-Elston regression"
author: "Martin Mullis"
date: "2025-04-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages
```{r}
#nessary for functions to work
library(parallel)
library(dplyr)

#used for visualization. not necessary for functions
library(scales)
```

Set working directory and load data
```{r}
# set working directory here. This should contain the phenotype file and the kinship file
wd <- "/PATH/TO/FILES/"

# load znorm phenotype data ('phenotypes_non_molecular'). 
#The first seven columns correspond to ID and covariates. 
#The remaining columns (7 onwards) correspond to phenotypes. 
load(paste0(wd,"DO_Phenotypes_non_molecular.RData"))
phenotypes <- phenotypes_non_molecular;rm(phenotypes_non_molecular) #rename for convenience

# load kinship data ('kin'). This is a normally list of 20 matrices that contain pairwise kinship...
# ...information among all mice in the data set for each of the 20 chromosomes.
# But in practice, only the first element of the list is actually used. 
# So from the github, I've uploaded a list that includes only the first chromosome. 
load(paste0(wd,"kinship_loco.RData"))
```

# Introduction
Haseman-Elston regression is a method for estimating the heritability (h2) of traits by regressing pairwise phenotype similarity scores against kinship for one or more sets of individuals. This method can also be extended to estimate the genetic correlation (rg) of pairs of traits via a structural equation model. The following functions constitute our implementation of Haseman-Elston for use with z-score normalized phenotype data from DO mice. 


# Z score normalization
znorm() takes in a vector of phenotype data and zscore-normalizes it. the resulting distribution will have a mean of 0, and a standard deviation and variance of 1.  
Arguments:  
**dist**: A vector of numerical phenotypes
```{r}
znorm <- function(dist){((dist - mean(dist))/sd(dist))}
```

Here's how it works:
```{r}
# simulate a normally distributed trait, measured in 1000 individuals...
# ...with a mean of 1 and a standard deviation of 3
n = 1000
mean = 1
sd = 3
trait <- rnorm(n=n,mean = mean,sd=sd)
print("pre-normalization")
print(paste0("mean = ",mean,"; sd = ",sd))

norm_trait <- znorm(trait)
print("post-normalizaton")
print(paste0("mean = ",mean(norm_trait),"; sd = ",sd(norm_trait)))

hist(trait,main="pre-normalization")
hist(norm_trait,main="post-normalization")
```

# Standard error correction
The se_correct() function is used for correcting the standard error (SE) of Haseman-Elston regression heritability (h2) and genetic correlation (rg) estimates. It implements an empirically derived correction factor that increases the SE a bit as sample size and h2 increase.  
Arguments: **se**, **h2**, and **n**.  
**se**: a standard error to be corrected.  
**h2**: the corresponding h2 estimate.  
**n**: the number of individuals used to estimate the h2.  
**output**: a corrected standard error estimate. 
```{r}
se_correct <- function(se,h2,n){
  factor <- sqrt(n) * (0.020225*h2 + 0.004225) + (-0.2352*h2 +0.9467)
  se_corrected <- se * factor
  return(se_corrected)
}
```

Demonstrate functionality of se_correct() by correcting a particular standard error (0.3) at different h2 levels and population sizes
```{r}
ns <- seq(100,1000,by=100) # set population sizes
se <- 0.3 # set a constant standard error

# get a bunch of corrected standard errors at three different h2 levels as a function of population size
ses <- data.frame(t(sapply(1:length(ns),function(x){
  low_h2 <- se_correct(se,h2=0.1,ns[x])
  med_h2 <- se_correct(se,h2=0.5,ns[x])
  high_h2 <- se_correct(se,h2=1.0,ns[x])
  out <- data.frame("n"=ns[x],"low_h2"=low_h2,"med_h2"=med_h2,"high_h2"=high_h2)
})))
print(ses)

plot1 <- plot(ses$n,ses$low_h2,col="royalblue",type="l",las=1,ylab="standard error",xlab="population size",ylim=c(0.2,0.5))
lines(ses$n,ses$med_h2,col="orchid2",type="l")
lines(ses$n,ses$high_h2,col="salmon",type="l")
legend("topleft",col=c("royalblue","orchid2","salmon"),legend=c("h2 = 0.1","h2 = 0.5","h2 = 1.0"),pch=16)
```

Note that as h2 and n grow, so too does the scale of the correction applied to the standard error. 

# Phenotype similarity scores & relatedness
The pairwise_product2() funciton takes in two sets of trait values and computes all combinations of pairwise products for the sets of values; these are used as phenotypic similarity scores in our implementation of Haseman-Elston regression. The function also takes in a kinship matrix so that it can assign pairwise kinship estimates to the similarity scores. The output is a dataframe that is ready for Haseman-Elston regression.  

'NA' values should be removed from the vectors of trait values prior to running this function. This is done automatically in the HEreg() function that calls pairwise_product2(). 

Arguments: **t1**, **t2**, **k**, **type**  
**t1:** Named vector of z-score normalized phenotype data. Names must correspond to Mouse ID and be present in the kinship matrix.  
**t2:** Second named vector of z-score normalized phenotype data.  
**k:** List of kinship matrices generated by the R/qtl2 package (Broman et al. 2019, 10.1534/genetics.118.301595).  
**type:** Specifies the type of Haseman-Elston regression to be performed.  
  - `"h"` corresponds to heritability.  
  - `"rg"` corresponds to genetic correlation.

The output is a four-column dataframe containing the following information:  
**id1**: ID of the first individual that contributed to the phenotypic similarity score.  
**id2**: ID of the second individual that contributed to the phenotypic similarity score.  
**prod**: The phenotypic similarity score (product of the z score normalized phenotypes).   
**relatedness**: The kinship of the two individuals.  
```{r}
pairwise_product2 <- function(t1,t2,k,type=c("h","rg")){
  kin <- k[[1]] # just need the kinship matrix for one autosome
  
  # make list of combinations to check
  ids1 <- names(t1)
  ids2 <- names(t2)
  
  kin_sub <- kin[rownames(kin) %in% ids1, colnames(kin) %in% ids2] # subset kinship matrix to only include individuals present in a sample of the data (this is written specifically for SE simulations).  When calculating cross-products normally, the two traits are already subset to only include phenotype information for individuals that are present in the kinship matrix. Thus, for normal h2 or GC calculations this line should not affect anything. 
  kin_sub <- kin_sub[match(names(t1),rownames(kin_sub)),]
  kin_sub <- kin_sub[,match(names(t2),colnames(kin_sub))]
  
  combs <- expand.grid(ids1,ids2)
  temp <- outer(t1,t2,FUN="*")
  if(type == "h"){
    ind <- which(upper.tri(temp,diag=F) , arr.ind = TRUE )
    
    out_df <- data.frame( col = dimnames(temp)[[2]][ind[,2]] ,
                          row = dimnames(temp)[[1]][ind[,1]] ,
                          prod = temp[ ind ] )
    
    ind2 <- which(upper.tri(kin_sub,diag=F) , arr.ind = TRUE )
    
    
    kin_sub2 <- data.frame( col = dimnames(kin_sub)[[2]][ind[,2]] ,
                            row = dimnames(kin_sub)[[1]][ind[,1]] ,
                            relatedness = kin_sub[ ind ] )
    
    out_df <- cbind(out_df,kin_sub2$relatedness)
    
  }else{
    out_df <-cbind(combs,as.vector(temp),as.vector(kin_sub))
  }
  
  colnames(out_df) <- c("id1","id2","prod","relatedness")
  out_df$id1 <- as.character(out_df$id1)
  out_df$id2 <- as.character(out_df$id2)
  out_df <- out_df[out_df$id1 != out_df$id2,] # exclude rows containing self-self comparisons
  
  rownames(out_df) <- 1:nrow(out_df)
  
  #end <- Sys.time()
  #end-start
  return(out_df)
}
```

Demonstrate funcitonality of pairwise_product2() by generating combinatorial products of either one phenotype (for h2 estimation) or two phentypes (for rg estimation). Set working directory path ('wd' above) to the phenotype data. 
```{r}
# sample first two phenotypes
traits <- colnames(phenotypes)[7:8]

# extract trait data and remove NAs
T1 <- phenotypes[,traits[1]]
names(T1) <- phenotypes$MouseID
T1 <- na.omit(T1)

T2 <- phenotypes[,traits[2]]
names(T2) <- phenotypes$MouseID
T2 <- na.omit(T2)

# compute combinatorial pairwise products and plot out for a single trait
pairwise_prod_h2 <- pairwise_product2(t1=T1,t2=T1,k=kin,type="h")
print(head(pairwise_prod_h2))
plot(pairwise_prod_h2$relatedness,pairwise_prod_h2$prod,col=alpha("black",0.3),pch=16,cex=0.4,las=1,xlab="kinship", ylab="phenotypic similarity")

pairwise_prod_rg <- pairwise_product2(t1=T1,t2=T2,k=kin,type="rg")
print(head(pairwise_prod_rg))
plot(pairwise_prod_rg$relatedness,pairwise_prod_rg$prod,col=alpha("black",0.3),pch=16,cex=0.4,las=1,xlab="kinship", ylab="phenotypic similarity")
```

# Haseman-Elston regression on z score normalized phenotype data
HEreg() performs Haseman-Elston regression on a set of traits and outputs relevant summary statistics. It uses phenotype data that has already been z-score normalized. It calls the pairwise_product2() and se_correct() functions above.

Arguments: **pheno**, **k**, **h2_df**,**type**,**traits**,**cores**,**correct**.  
**pheno**: A dataframe consisting of z-score normalized phenotype values. The first column should be titled "MouseID" and contain ID information for the animals that corresponds to the row and column names of the kinship matrix. Subsequent columns correspond to normalized phenotype data.  
**k**: a list of kinship matrices of the sort generated by the R/qtl2 package (Broman et al. 2019, 10.1534/genetics.118.301595).   
**h2_df**: only necessary if type = "gc". this is a dataframe of heritability values generated by HEreg(). These h2 values are used in the structural equation model to estimate genetic correlations for pairs of traits.  
**traits**: an n-by-2 matrix providing combinations of traits to use in the analysis.  
**cores**: the number of cores to use for parallelization. be mindful of memory constraints.  
**correct**: implement SE correction?  

**Output**: an n-by-14 dataframe.  
**trait1**: The first trait used in the estimation of rg. Will be the same as **trait2** if type="h".    
**trait2**: The second trait used in the estimation of rg. Will be the same as **trait1** if type="h".  
**gcor**: The genetic correlation estimate for the pair of traits. Will be **1** if type="h" as traits are always perfectly genetically correlated with themselves. For type="gc", this may be **NA** if the h2 of one of the traits is too low (e.g. h2 < 0).  
**gcor_p**: The p-value associated with the rg estimate.  
**pcor**: The (Pearson) phenotypic correlation of the pair of traits. Will be **1** if type="h". Will be **NA** if a pair of traits were measured in a nonoverlapping set of animals.  
**pcor_p**: The p-value associated with the pcor value.  
**h2**: If type="h", this is the heritability of the trait. If type = "gc", this is **NA**.  
**se_gc**: The standard error of the genetic correlation estimate.  
**se_h2**: The standard error of the h2 estimate. Will be **NA** if type="gc". 
**ob**: The regression coefficient of Haseman-Elston regression, divided by 2. This is an intermediate value used initially to assess the performance of the function. Not really useful.  
**h2_trait1**: The heritability of **trait1**.  
**h2_trait2**: The heritability of **trait2**.  
**n_1**: The number of individuals from **trait1** used in the regression.  
**n_2**: The number of individuals from **trait2** used in the regression.  
```{r}
HEreg <- function(pheno,k,h2_df,type=c("h","gc"),traits,cores,correct=T){
  names <- colnames(pheno)[2:ncol(pheno)]
  if(is.null(traits)){
    if(type == "h"){
      pairs <- t(sapply(1:length(names),function(x){rep(names[x],times=2)}))
    }else{
      herit_df <- h2_df
      pairs <- t(combn(names,2))
    }
  }else{
    pairs <- traits
    herit_df <- h2_df
  }
  
  cor_df <- mclapply(1:nrow(pairs),function(x){
    if(x %% 1000 == 0){print(x)}
    start <- Sys.time()
    # retreive names of traits used for HE regression
    trait1 <- pairs[x,1]
    trait2 <- pairs[x,2]
    
    # access trait values from phenotype dataframe
    T1 <- pheno[is.na(pheno[,trait1]) == F,trait1]
    names(T1) <- pheno[is.na(pheno[,trait1]) == F,1]
    T2 <- pheno[is.na(pheno[,trait2]) == F,trait2]
    names(T2) <- pheno[is.na(pheno[,trait2]) == F,1]
    
    n_1 <- length(T1)
    n_2 <- length(T2)
    
    # there are no non-na values, return na for the output
    if(length(T1) == 0 | length(T2) == 0){
      out <- c(trait1,trait2,NA,NA,NA,NA,NA,NA,NA,NA,NA,NA,n_1,n_2)
      return(out)
    }
    
    # subset to only include individuals in the kinship matrix
    T1 <- T1[which(names(T1) %in% colnames(k[[1]]))]
    T2 <- T2[which(names(T2) %in% colnames(k[[1]]))]
    
    # pull shared values to calculate phenotypic correlation
    shared <- intersect(names(T1),names(T2))
    if(length(shared) >= 3){
      T1_sub <- T1[`shared`]
      T2_sub <- T2[`shared`]
      pcor <- cor(T1_sub,T2_sub)
      pcor_p <- cor.test(T1_sub,T2_sub)$p.value
    }else{
      pcor <- NA
      pcor_p <- NA
    }
    
    # get trait product and relatedness for each pair of animals
    data <- pairwise_product2(t1=T1,t2=T2,k=k,type=type)
    
    fit <- lm(data$prod ~ data$relatedness) 
    gcor_p <- summary(fit)$coefficients[2,4]
    b <- fit$coefficients[2]
    
    ob <- b/(2*var(T1)) # because all traits are z-normalized in advance, the variance of both traits is going to be 1, so it doesn't matter which trait is used. 
    
    if(type == "h"){
      h2 <- ob
      h2_trait1 <- h2
      h2_trait2 <- h2
    }else{
      h2 <- NA
      h2_trait1 <- herit_df[herit_df$trait1 == trait1,"h2"]
      h2_trait2 <- herit_df[herit_df$trait1 == trait2,"h2"]
      se_h2_trait1 <- herit_df[herit_df$trait1 == trait1,"se_h2"]
      se_h2_trait2 <- herit_df[herit_df$trait1 == trait2,"se_h2"]
    }
    
    gcor <- ob/(sqrt(h2_trait1)*sqrt(h2_trait2))
    
    if(type == "gc"){
      b_error <- sqrt(sum(((data$prod - mean(data$prod))^2)) / (nrow(data) - 2)) / sqrt(sum(((data$relatedness - mean(data$relatedness))^2)))
      
      ob_error <- b_error/(2*var(T1))
      
      if(correct==T){
        ob_error <- se_correct(n=sqrt(n_1*n_2),h2=ob,se=ob_error)
      }else{
        ob_error <- ob_error
      }
      
      
      # source: http://www.met.rdg.ac.uk/~swrhgnrj/combining_errors.pdf
      se_gc <- sqrt( (ob_error/ob)^2 + ((.5*sqrt(h2_trait1)*se_h2_trait1) / sqrt(h2_trait1))^2 + ((.5*sqrt(se_h2_trait2)*se_h2_trait2) / sqrt(h2_trait2))^2 )
      
      se_h2 <- NA
    }else{
      b_error <- sqrt(sum(((data$prod - mean(data$prod))^2)) / (nrow(data) - 2)) / sqrt(sum(((data$relatedness - mean(data$relatedness))^2)))
      
      ob_error <- b_error/(2*var(T1))
      
      se_h2 <- ob_error
      
      if(correct==T){
        se_h2 <- se_correct(n=n_1,h2=h2,se=se_h2)
      }else{
        se_h2 <- se_h2
      }

      # source: http://www.met.rdg.ac.uk/~swrhgnrj/combining_errors.pdf
      # https://www.met.reading.ac.uk/~swrhgnrj/
      # https://scholar.google.com/citations?hl=en&user=-vUyGWsAAAAJ&view_op=list_works&sortby=pubdate
      se_gc <- sqrt( (ob_error/ob)^2 + ((.5*sqrt(h2)*se_h2) / sqrt(h2))^2 + ((.5*sqrt(se_h2)*se_h2) / sqrt(h2))^2)
    }
    out <- c(trait1,trait2,gcor,gcor_p,pcor,pcor_p,h2,se_gc,se_h2,ob,h2_trait1,h2_trait2,n_1,n_2)
    
    end <- Sys.time()
    #print(end-start)
    return(out)
  },mc.cores=cores)
  
  cor_df <- do.call('rbind',cor_df)
  
  colnames(cor_df) <- c("trait1","trait2","gcor","gcor_p","pcor","pcor_p","h2","se_gc","se_h2","ob","h2_trait1","h2_trait2","n_1","n_2")
  
  cor_df <- data.frame(cor_df)
  for(i in 3:ncol(cor_df)){
    cor_df[,i] <- as.numeric(cor_df[,i])
  }
  for(i in 3:(ncol(cor_df)-2)){
    cor_df[,i] <- round(cor_df[,i],3)
  }
  return(cor_df)
}
```

Demonstrate utility of HEreg() function by estimating heritability for a subset of phenotypes 
```{r}
#discard covariate data in the phenotypes dataframe
phenotypes <- phenotypes[,c(1,7:ncol(phenotypes))]

#specify a subset of phenotypes to test the function on
traits <- cbind(colnames(phenotypes[,2:20]),colnames(phenotypes[,2:20]))
print("Correct format for 'traits' matrix for h2 estimation:")
print(head(traits))

# compute h2 for the first 20 traits in our dataset
h2 <- HEreg(pheno=phenotypes,h2_df=NULL,k=kin,type="h",traits=traits,cores=2,correct=T)
print("Heritability of traits:")
print(head(h2))

# subset our focal traits to those that are heritable. HEreg can't produce valid estimates if the HE or SE are really big. 
traits_subset <- h2$trait1[h2$h2 > 0.1]

# specify combinations of traits to compute rg for
traits2 <- t(combn(traits,2))
print("Correct format for 'traits' matrix for rg estimation:")
print(head(traits2))

# compute rg among the first 20 traits in our dataset
rg <- HEreg(pheno=phenotypes,h2_df=h2,k=kin,type="gc",traits=traits2,cores=2,correct=T)
print("Genetic correlations among heritable traits:")
head(rg)
```
